---
title: "TreeBUGS: Hierarchical MPT Modeling"
author: "Daniel W. Heck, Nina Arnold, & Denis Arnold"
date: "`r Sys.Date()`"
number_section: true
output:
      rmarkdown::html_vignette:
      toc: true
vignette: >
  %\VignetteIndexEntry{TreeBUGS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## General Procedure of Using TreeBUGS

In the most simple user scenario, the following steps are required:

1. Define path to existing MPT model file in .eqn format (cf. multiTree; Moshagen, 2010)
2. Define path to data set with individual frequencies (.csv file: comma separated, rows=persons, columns=labeled categories)
3. Call one of the fitting functions `betaMPT` or `traitMPT` (exact code below)
4. Check convergence of MCMC chains
5. Summarize and plot results

In the following, these steps are explained in more detail. Note that TreeBUGS requires a valid and recent version of the software JAGS (http://mcmc-jags.sourceforge.net/).



### 1. Step: MPT model file in EQN syntax

The model needs to be passed in the standard .eqn file format (e.g., as in multiTree; Moshagen, 2010). As an example, consider the most simple two-high-threshold model (2HTM), each line defines a single processing path containing tree label, category label, and model equations:

```
# Title (ignored): 2HTM
Target    Hit    Do
Target    Hit    (1-Do)*g
Target    Miss   (1-Do)*(1-g)
Lure      FA     (1-Dn)*g
Lure      CR     (1-Dn)*(1-g)
Lure      CR     (1-Dn)
``` 

Note that model equations require the multiplication sign `*` and parameters should not be summarized, e.g., by `a^2*(1-a)`. As an input for TreeBUGS, the model file (e.g., `"2htm.txt"`) needs to be saved in the current working directory. Otherwise, the relative or absolute path to the file must be specified (e.g., `/models/2htm.txt`). To check how TreeBUGS interprets a given .eqn-file, use:
```{r, eval=F}
readEQN(file = "pathToFile.eqn",       # relative or absolute path 
        restrictions = list("Dn=Do"),  # equality constraints
        paramOrder = TRUE)             # show parameter order
```

Equality restrictions on the MPT parameters can either be provided in a list:
```{r, eval=FALSE}
restrictions = list("Dn=Do", "g=0.5")
```
or by a path to a text file on the hard drive (e.g., `restrictions="pathToFile.txt"`) that contains the equality constraints:
```
Dn=Do
g=0.5
```



### 2. Step: Data Set with Individual Frequencies

As default, the data are provided in a comma-separated text file (.csv) in the following format:

```
Hit,   Miss,     FA,    CR
20,      10,      5,    25
13,       7,      9,    21 
15,       5,      6,    14
.....

```

Note that the first line contains the category labels, which must match the category labels from the .eqn-file. The remaining rows contain individual frequencies. Similarly as for the .eqn file, the path to the data file can either be specified as  `data_ind.csv` if it is in the current working directory, or as an relative or absolute path (e.g., `"C:/models/data_ind.csv"`). 

Alternatively, a data.frame or matrix with appropriate column names (i.e., category labels) can be provided.



### 3. Step: Fit Hierarchical MPT Model

An hierarchical **Beta-MPT model** is fitted with the following code:

```{r, eval=FALSE}
# load the package:
require("TreeBUGS")

# fit the model:
fitHierarchicalMPT <- betaMPT(eqnfile="2htm.txt",              # .eqn file
                              data="data_ind.csv",             # individual data
                              restrictions=c("do=dn"),         # parameter restrictions (or path to file)
                              
                              ### optional MCMC input:                 
                              n.iter = 50000,                  # number of iterations
                              n.burnin = 5000,                 # number of burnin samples that are removed
                              n.thin = 5,                      # thinning rate of removing samples
                              n.chains = 3                     # number of MCMC chains (run in parallel)
)

```

Similarly, a latent-trait model is fitted simply by replacing `betaMPT` by `traitMPT`.


### 4. Step: Check convergence of MCMC chains

The functions `betaMPT` and `traitMPT` return a list including the original samples from the MCMC sampler for convergence checks. The MCMC samples are stored in `fittedModel$runjags$mcmc` as an `mcmc.list` object (see the package \link{coda} for convergence diagniostics). TreeBUGS provides a handy wrapper to access the most important plotting functions:

```{r, eval=FALSE}
# Default: Traceplot and density
plot(fittedModel,    # fitted model
     parameter ="mean"            # which parameter to plot
)
# further arguments are passed to ?plot.mcmc.list

# Auto-correlation plots:
plot(fittedModel,  parameter = "mean", type = "acf")

# Gelman-Rubin plots:
plot(fittedModel,  parameter = "mean", type = "gelman")
```

See \link{coda} and \link{runjags} for further convergence statistics and plots. Note that inferences from from the model are likely not to be valid if the Markov-Chain Monte-Carlo (MCMC) sampler did not reach convergence!

### 5. Step: Summarize and Plot Model

TreeBUGS produces an MPT-tailored summary of parameter estimates and convergence statistics:

* Information about the parameter posterior distribution: Mean, SD, Median, 2.5% and 97.5% quantiles
* Convergence: R²-Statistic (should be close to 1.00 for all parameters) and number of effective samples when accounting for auto-correlation (should be large)

To obtain the summary after fitting the model, simly use:
```{r, eval=FALSE}
summary(fitHierarchicalMPT)
```

The following functions allow to plot parameter estimates, distributions, goodness of fit, and raw frequencies:

```{r, eval=FALSE}
plotParam(fitHierarchicalMPT,     # fitted model
          includeIndividual=TRUE  # whether to plot individual estimates
)
plotDistribution(fittedModel, ...)   # plot estimated hierarchical parameter distribution
plotFit(fittedModel,...)             # plot observed and predicted mean frequencies
plotFreq(fittedModel,...)            # plot individual and mean raw frequencies per tree

```

Parameter estimates (posterior mean, median, SD) can be extracted by using:

```{r, eval=FALSE}
getParam(fitHierarchicalMPT, parameter="theta", posterior="mean")
```


## Extended Modeling Possibilities

### A) Including Continuous Covariates

It is possible to sample and estimate correlations of continuous covariates (e.g., age) with the individual MPT parameters. Note that this does not influence the model estimates (the estimated MPT parameters are only used repeatedly to compute a correlation). In contrast, in the latent-trait MPT model, variables can also be included as predictors to account for interindividual variance in MPT parameters, which influences the parameter estimates:
$$\theta_{is} = \Phi(\mu_s + \delta_i + \gamma_i X_i)$$

The following arguments are used to specify the desired covariance structure:

* `covData`: Either a data frame or the path to a csv data file separated by commas `,` hat contains the covariates
    * Rows: The order of individuals must be identical to the order in the frequency data (`data`)
    * Columns: Covariates must habe column names different from the parameters
* `covStructure`: Which correlations should be sampled? Either a list or path to a text file in which the assignment of MPT parameters to covariates is coded as follows:
    * `one or more MPT parameter labels  ; one or more covariate labels `  (separated by a semicolon)
    * For instance: `list("Do Dn ; IQ",  "g ; age extraversion")` 
    * Multiple combinations are included by separate entries in the list or by separate lines in the text file (redundant combinations are removed automatically)
* `corProbit`: whether to correlate MPT parameters on the probability scale (default for beta MPT) or on the latent probit scale (default for latent-trait MPT)
* `predStructure`: Which MPT parameters to predict by which variables? (only for latent-trait model; similar syntax as `covStructure`)
    
Overall, the code could look like this:

```{r, eval=F}
fitMPT <- traitMPT(eqnfile = "2htm.txt", 
                   data = "data_ind.csv",   
                   restrictions = c("do=dn"),  
                   covData = "data_covariates.csv", 
                   covStructure = list("Do Dn ; weight",  "g ; age height"),  # correlations
                   predStructure = list("Do Dn ; IQ"),   # IQ as predictor variable
                   ...
```

After fitting the model, the results appear when typing `summary(fitMPT)`.


### B) Including Discrete Factors in a Latent-Trait MPT Model

In the latent-trait model, it is possible to include discrete factors as predictor variables, similar as in the general linear model formulation of an ANOVA. Compared to continuous covariates only the following changes:

* New argument `predType`, which is a character vector that assignes each column in `covData` a specific tyoe (i.e., how it is used in `predStructure`). Specifically, predictor variables can be set as
    * continuous (`"c"`)
    * fixed effect (`"f"`)
    * random effect (`"r"`)
* `covData` can have columns with character or factor variables (numeric columns can be specified as factors using `predType`)
* By default, character variables in `covData` are included as fixed effects
* The order of `predType` has to match the column order of `covData`

Note that the same parameter covariance structure is assumed in each group, which might produce more reliable estimates than specifying a separate MPT tree for each condition (and thus assuming a separate parameter covariance matrix in each group). An example might be: 

```{r, eval=F}
fitMPT <- traitMPT(eqnfile = "2htm.txt", 
                   data = "data_ind.csv",   
                   covData = "data_covariates.csv", 
                   predStructure = list("Do ; factor1", "Dn ; factor2"),  #predictor
                   predType = c("c", "c", "f", "r")
                   ...
```

Group estimates can be obtained by 

```{r, eval=F}
getGroupMeans(traitMPT)
```


Multiple factors can be included, but currently it is not possible to include interactions. For an introduction to Bayesian ANOVA, see Rouder et al., 2012.


### C) Sampling Transformed Parameters

The argument `transformedParameters`allows to sample any mathematical function of the actual MPT parameters. This is helpful to test differences between two core MPT parameters or obtain reparameterized versions of the parameters (e.g., for order constraints). For instance, the difference between two MPT parameters can be computed using

```{r, eval=FALSE}
transformedParameters = list("deltaG = G_1 - G_2",     # difference of parameters
                             "G1_larger = G_1 > G_2")  # indicator variable (testing order constraints)
```

If the parameters are different, the 95% posterior interval of the parameter `deltaG` should exclude zero. 

Transformed parameters are also helpful if the model contains reparameterizations of order constraints. For instance, if $a<b$ is replaced by $a = s_a * b$ (the standard procedure in multiTree), the EQN file includes the parameters `b` and  `s_a`, but the interest is in `a`, which can be obtained by `transformedParameters = list("a = s_a * b")` . 

Note the following about the correct specification of transformed parameters:

* transformed parameters require new, unique labels left of the equality sign `=` 
* parameters on the right hand must match with the MPT parameters in the .eqn-file
* transformed parameters are computed and monitored on the group-level only

### D) Generate Simulated Data Sets

Simulated data sets are in general useful to check the robustness of the estimators and the sample size requirements. TreeBUGS includes functions to generate data sets of individual frequencies for both the Beta-MPT and the latent-trait MPT model. 

```{r, eval=FALSE}
# beta-MPT
genBeta <- genBetaMPT(N = 100,                             # number of participants     
                      numItems = c(Target=250, Lure=250),  # number of responses per tree
                      eqnfile = "2htm.eqn",                # path to MPT file
                      mean = c(Do=.7, Dn=.7, g=.5),        # true group-level parameters
                      sd =   c(Do=.1, Dn=.1, g=.05))       # SD of individual parameters
                   
# latent-trait MPT
genTrait <- genTraitMPT(N = 100,                             # number of participants     
                        numItems = c(Target=250, Lure=250),  # number of responses per tree
                        eqnfile = "2htm.eqn",                # path to MPT file
                        mean = c(Do=.7, Dn=.7, g=.5),        # true group-level parameters
                        sigma =   c(Do=.25, Dn=.25, g=.05),  # SD of latent (!) individual parameters
                        rho = diag(3))                       # correlation matrix. here: no correlation
```

The resulting data sets contain both the generated frequncies (`genTrait$data`) and the data-generating group and individual parameters (`genTrait$parameters` )



## References
 
* Erdfelder, E., Auer, T.-S., Hilbig, B. E., Assfalg, A., Moshagen, M., & Nadarevic, L. (2009). Multinomial processing tree models: A review of the literature. Journal of Psychology, 217, 108–124. http://doi.org/10.1027/0044-3409.217.3.108

* Klauer, K. C. (2010). Hierarchical multinomial processing tree models: A latent-trait approach. Psychometrika, 75, 70–98. http://doi.org/10.1007/s11336-009-9141-0

* Matzke, D., Dolan, C. V., Batchelder, W. H., & Wagenmakers, E.-J. (2015). Bayesian estimation of multinomial processing tree models with heterogeneity in participants and items. Psychometrika, 80, 205–235.

* Moshagen, M. (2010). multiTree: A computer program for the analysis of multinomial processing tree models. Behavior Research Methods, 42, 42–54. http://doi.org/10.3758/BRM.42.1.42

* Smith, J. B., & Batchelder, W. H. (2010). Beta-MPT: Multinomial processing tree models for addressing individual differences. Journal of Mathematical Psychology, 54, 167–183. http://doi.org/10.1016/j.jmp.2009.06.007

